Rice yield improvements through plant breeding are offset by inherent yield declines over time

### 2017-10-29 output/breed_paper.Rnw
Updated with Bruce's comments - need to revise.

Constraints on arable land are increasing simultaneous with the need
to increase total food production to meet a growing demand
\citep{Foley2011a, Godfray2010, Mueller2012}, which has necessitated
harvesting more grain per unit land area \citep{Lobell2009,
Tittonell2014}. Historically, agricultural research has been
successful in staving off the ``Malthusian catastrophe'' of demand
surpassing supply via continued yield improvement of staple
crops. However, many production systems are experiencing plateaus in grain
yield \citep{Grassini2013, Peng2010}. If this trend continues,
improvements per unit land area are no longer possible, and an
increase in the area under cultivation will be needed to meet food
demand, which carries with it undesirable ecological
implications~\citep{Foley2011a}. Therefore, it is of critical importance to better
understand why the rate of increase in grain yields has declined or
leveled off in intensified production systems.

California (CA) rice represents one such production system. Rice is
grown primarily in the Sacramento Valley which is characterized by
having a Mediterranean climate, with long days, dry growing season,
and relatively free of pests and diseases, which lead to some of the
highest yields in the world~\citep{FAO2016}. Production in CA is
predominately focused on premium quality medium grain
\textit{japonica} cultivars (e.g., CalRose rice), and rice from CA is
recognized globally for its quality
(\url{http://agfax.com/2015/11/03/rice-calrose-wins-best-rice-world-competition/}). Most
cultivars in use in CA are developed by the California Cooperative
Rice Research Foundation (CCRRF; a collaboration between the
University of California, the USDA-Agricultural Research Service, and
farmer-funded research). In part due to improvements in rice genetics,
rice yields in CA increased rapidly during the period from 1920 --
1990; however, since the 1990s, the rate of yield increase has slowed
(Fig. \ref{fig:state_yield}) despite continuous crop
improvement. Based on a yield-gap analysis, on-farm yields in the
major rice growing region of CA is 73 to 76\% of maximum yield
potential~\citep{espe16a}. In highly intensive systems such as this,
\citet{grassini2011} have shown that farmers are capable of attaining
85\% of the maximum yield potential. Thus, given that farmers are near
the attainable yield potential, increases in yield are likely to be
slow and \citet{espe16a} reported that indeed this was the case
with yields on average increasing by about 50 to 62 kg ha-1 between
1999 and 2014.

Certainly one question in looking at this situation is, “do new
varieties have increased yield potential”? Like many breeding
programs, one of the goals of the CCRRF is to increase yields through
breeding varieties with increased yield potential. However, when the
popular cultivar M-202 was
released in 1985, it was reported to yield an average of 11 Mg
ha\tsups{-1} \citep[14\% moisture content (MC);][]{m202}. In 2015, the
reported average yield of the newest released cultivar M-209 was
10.8 Mg ha\tsups{-1} \citep[14\% MC;][]{m209}. However, in annual
breeding reports and varietal release announcements, based on yield
performance in side-by-side variety trails, data show that new
varieties usually out yield the older varieties. In other rice
systems, a decrease in a cultivar's yield performance over time has
been observed \citep{Peng1998, Peng2010, Peng2000, Datta1995}, often
termed yield erosion \citep{Peng1998}. It has been speculated that
this yield erosion is due to the inability of a cultivar to adapt to
changing biotic and abiotic conditions~\citep{Peng1998, Peng2010,
Peng2000, Datta1995}. Yield erosion may explain the apparent
contradiction above, in which side-by-side yield comparisons show
yield improvements in varieties while overall there may be little
change in yield potential. To address this issue we teseted the
hypothesis that if rice cultivar yields erode over time then apparent
yield increases in side-by-side yield comparison tests will not
reflect true yield increases.

@@

As part of the annual Statewide Variety Trials, newly developed
entries are tested against officially released cultivars (checks) at
the RES in several trials spanning multiple planting dates per
year. For the purposes of this study, we concentrated on released
medium grain cultivars, as these cultivars constitute approximately
90\% of the planted rice area in CA. Newly developed entries are typcically
evaluated in these trials for 3 to 5 years prior to
their official release. For these cultivars planted in breeding trials
at the RES from \Sexpr{min(mod_data[,'year'])} --
\Sexpr{max(mod_data[,'year'])} plot level observations (replicated
3--4 times in a CRB design) were collected leading to a final data set
of \Sexpr{format(nrow(mod_data), big.mark = ',')} observations
representing \Sexpr{length(unique(mod_data[,'id']))} cultivars and
included every publicly released medium grain cultivar developed by
the CCRRF from \Sexpr{min(mod_data[,'release_yr'])} --
\Sexpr{max(mod_data[,'release_yr'])}. For each cultivar, the official
year of public release and the time elapsed (in years) since release was
determined.

@@

\subsection{Genetic yield improvements}

There was agreement between the estimates from the Bayesian and
classical approaches. The Bayesian estimates are presented here,
though the classical estimates can be found in the supplemental
material. Comparing the cultivars' yield during each's initial time
(average 3 years) in the experimental trials confirmed that genetic
yield improvements have been consistant over the 30 year study period
(Fig. \ref{fig:first_yrs}). As estimated by the model including both
yield erosion and yield improvement over time, there is no conclusive
evidence that the yields of medium grain rice cultivars selected and
released by CCRRF's breeding program have, on average, increased over
the last 30 years (Table \ref{tbl:coef}). The full model which
corrected for both the year of release and yield erosion was estimated
to have better out-of-sample predictive accuracy compared to reduced
models containing only one predictor (See supplemental material for
further details). The estimated change in grain yield per year in the
CCRRF's breeding trials at the RES was estimated to be a negligible
\Sexpr{release_yr_coef} kg ha\tsups{-1} year\tsups{-1}, with both
small positive and negative trends within the credible interval (90\%
CI: \Sexpr{rel_CI[1]} to \Sexpr{rel_CI[2]})
(Fig. \ref{fig:first_yrs}).

When yield erosion was not corrected for in the analysis (i.e.,
omitted from the model), there is an apparent small increase per year
(approximately \Sexpr{only_release_coef} kg ha\tsups{-1}
year\tsups{-1}; 90\% CI \Sexpr{only_release_CI[1]} to
\Sexpr{only_release_CI[2]}) (Fig. \ref{fig:wo_erosion}). Although this
estimate is flawed (i.e., it ignores the confounding effect of yield
erosion over time and performed worse in out-of-sample predictive
accuracy), it confirms previous analyses of side-by-side comparisons
which concluded that newer cultivars out-preform older ones.

After correcting for yield erosion, some cultivars were higher
yielding compared to the overall average (as estimated by random
intercepts). The marginal estimate for the performance of a cultivar
averaged over the uncertainty in the other parameters in the model
(i.e., the random effect) shows three cultivars with higher than
average yields:
\Sexpr{paste(gsub('M','M-',rev(row.names(id_est))[c(1,2,4)]),collapse=',')}
(Figs. \ref{fig:first_yrs} \& \ref{fig:ids_comp}). A fourth, M-209,
might also have higher than average yield, though the CI for M-209
overlaps with zero (Fig. \ref{fig:ids_comp}).

@@

@
\caption{The relationship between yield and the years since release
  for medium grain rice cultivars. The estimated erosion of yield over time
  (averaged over differences in cultivar and year) is estimated to be
  \Sexpr{yrs_coef} kg ha\tsups{-1} year\tsups{-1}. Points ($\bullet$)
  and bars (|) represent the mean and standard error of the raw data,
  respectively. The shaded area is the 90\% credible interval.}
% For each year, point size by n (open circles)
\label{fig:RES_year}
\end{figure}

@
\caption{The performance of two major cultivars over the period
\Sexpr{min(mod_data[,'year'])} -- \Sexpr{max(mod_data[,'year'])}. The
solid line is the estimated yield erosion (\Sexpr{yrs_coef} kg
ha\tsups{-1} year\tsups{-1}), while the dashed line is the published
3-year average yield of M-202 at its release (1982-1984). Both
cultivars have a similar yield at their respective year of release,
and similar decreases over time.}
\label{fig:compare_yrs}
\end{figure}

The average decrease in yield over time once a cultivar is selected
(i.e., yield erosion) was estimated as \Sexpr{yrs_coef} kg
ha\tsups{-1} year\tsups{-1} (90\% CI: \Sexpr{RES_CI[2]} to
\Sexpr{RES_CI[1]}). The oldest cultivar in the trials had lost an
average of \Sexpr{round(yrs_coef*-36/1000,1)} Mg ha\tsups{-1} since
its release. The two most dominate medium grain rice cultivars, M-202
(widely planted from the late 1980s until the early 2000s) and M-206
(currently the most widely planted), experienced
similar trends in yield decline (Fig. \ref{fig:compare_yrs}; Table
\ref{tbl:coef}). The currently estimated
\Sexpr{m202_v_206} kg ha\tsups{-1} yield advantage of M-206 over M-202
can be attributed solely to the erosion of M-202's yield over time
(Figs \ref{fig:first_yrs} \& \ref{fig:compare_yrs}). Further
supporting this estimate of yield erosion, both cultivars yielded similarly at their
respective year of release (Fig. \ref{fig:compare_yrs}), despite the
fact that M-202 was released 19 years prior to M-206
(Fig. \ref{fig:compare_yrs}, Table \ref{tbl:ids}).

@@

We do not find conclusive evidence of increased genetic yield
potential in medium grain rice cultivars released by CCRRF's breeding
improvement program over the period \Sexpr{min(mod_data[,'year'])} to
\Sexpr{max(mod_data[,'year'])}. Since the estimated yield improvement
spans from \Sexpr{rel_CI[1]} to \Sexpr{rel_CI[2]} kg ha\tsups{-1}
year\tsups{-1}, both increases and decreases in the genetic yield
potential over the study period are plausible. However, even the
estimated yield improvement that does not account for yield erosion
(\Sexpr{only_release_coef} kg ha\tsups{-1} year\tsups{-1};
Fig. \ref{fig:wo_erosion}) is much lower than previous estimated,
i.e., 37 to 60 kg ha\tsups{-1} year\tsups{-1} \citep{Omar2014}. Based on the
estimates from this study, there is less than a 1\% probability that
the true value is higher than 45 kg ha\tsups{-1}
year\tsups{-1}. Therefore, we conclude that the CCRRF breeding program
has likely maintained but not substantially improved the yield
potential in the face of changing biotic and abiotic pressures.

We estimate that yield erosion has decreased the yield of medium grain
rice cultivars released by the CCRRF over the last 30+ years by an
average of \Sexpr{yrs_coef} Mg ha\tsups{-1}, potentially offsetting
any gains made by the crop improvement program. Biotic and abiotic forces have
previously been speculated to be responsible for yield erosion, since
the cultivar's genetics remain stable from the time of selection but
the conditions the cultivar may experience over time are continuously
changing \citep{Peng1998, Peng2000, Peng2010}. The trend observed here
is consistent with previous reports of the erosion of genetic gains of
cultivar IR8 at IRRI \citep{Peng1998, Peng2000}, suggesting the
erosion of yield over time is a phenomenon that widely affects rice
breeding programs. This observation highlights both the need for maintenance
of plant breeding and the difficulty of increasing genetic yield potential,
especially in systems such as CA where yield is approaching the
physiological limit \citep{espe16a}. Without maintenance breeding
(i.e., continuously selecting cultivars well adapted to current
conditions), CA rice production systems are predicted to have lost an
estimated \Sexpr{round(yrs_coef*(-36/1000),1)} Mg ha\tsups{-1} grain
yield over the study period.

However, a goal of many breeding programs is to not only maintain the
benchmark yield of the previous ``standard'' but to surpass it,
especially if we are to meet the target of 50\% increased food
production by 2050 \citep{Foley2011a}. Given the estimated yield
erosion, to hit an ambitious target of 0.5 Mg ha\tsups{-1}
(approximately 5\%) increase in yield per release (with roughly 4 to 5
years between releases), a new cultivar needs to yield an additional
\Sexpr{round(yrs_coef*-4,-1)} to \Sexpr{round(yrs_coef*-5,-1)} kg
ha\tsups{-1} in side-by-side comparison in addition to the 0.5 Mg
ha\tsups{-1} target improvement. Often side-by-side yield improvements
show an improvement in grain yield of 2 to 3\% per release, in which
case no absolute improvements in yield have been made beyond
maintaining yield. The evidence here is consistent with the
observation that the genetic yield potential has not increased
substantially in the last 30+ years (Fig. \ref{fig:first_yrs}).

@@

California state average yields have increased slowly
despite there being little to no gains in the breeding trials
(Fig. \ref{fig:state_yield}). The evidence here suggests mechanisms
other than increased genetic yield potential are responsible for these gains
(Fig. \ref{fig:compare_yrs}). Improvement in agronomic management,
including more efficient nutrient management \citep{Lundy2012,
Lundy2015, linquist2009}, weed management \citep{Saito2008,
Pittelkow2012, linquist2008}, and other factors may all be contributing to
increases in the state average yield. However, these gains are far less
than those experienced during the preceeding era, and by their
nature agronomic gains cannot surpass the genetic yield potential
\citep{Fischer2015}. Judging from yields achieved at the RES
(Figs. \ref{fig:RES_year} \& \ref{fig:compare_yrs}), a
researcher-managed production system implementing best management
practices in more controlled situations compared to farmers' fields, state
average yields (approximately 10 Mg ha\tsups{-1} at the time of
writing; Fig. \ref{fig:state_yield}) may be approaching the current
genetic limit. Assuming that the level of management possible in small
experimental plots cannot be attained in farmers' fields and hence
experimental yields represent the attainable yield ceiling
\citep{Fischer2015}, agronomic gains in farmer's fields are nearing
their potential, consistent with physiological model-based estimates
of the attainable yield gap in CA rice production systems
\citep{espe16a}.

There are several cultivars that were found to have better than
average yields, even after correcting for yield erosion. However,
these cultivars (M-201, M-204, M-205, and possibly M-209;
Fig. \ref{fig:ids_comp})) are not widely adapted. It is known that
cool temperatures during the booting stage can have a large impact on
grain yields \citep{espe2017}, and the RES
experiences warmer temperatures than many other rice growing areas in
CA \citep{espe2017}. These cultivars may perform well at the RES but
not widely in CA. Therefore, it should be noted that while our
investigation has been concerned about increasing yield potential, the
estimates here might not capture yield stability. Yield stability is
notoriously difficult to quantify \citep[e.g.,][]{lobell2011}, hence
further investigation is needed to explore how yield stability has
changed over the study period.

@@

Planning to meet future
demand necessitates increases in grain yield \citep{Godfray2010,
Foley2011a, Mueller2012}. Meeting the challenge to produce more grain
without expansion of land under cultivation will rely on new genetics
with increased yield potential. The evidence here suggests that to
accomplish true increases over time, the improvement target needs take
into account decreases in the current ``standard'' cultivars' yield
over time (Fig. \ref{fig:theory}). While side-by-side comparisons will
continue to be useful, comparisons across years to each cultivar's
benchmark yield at the time of release should provide information on
the absolute improvements being made in yield potential over time. New
technologies may need to be adapted to further increase yield
potential, including hybrid rice.

@@

Improvement in grain yield is essential to meeting the challenges of
feeding a growing population in a changing world. To achieve these
improvements, we discuss here that (1) continuous maintenance via plant
improvement is required, without which yields would decrease over
time, (2) the challenge of improving genetic yield potential is
substantial, since absolute gains require improvement in yield in
excess of the maintenance threshold, and (3) increasing yield through
improved management (agronomy) cannot exceed the ceiling of genetic
yield potential. Taken together, these conclusions underscore both the
necessity and substantial challenges facing rice breeding. We suggest
here that historical benchmarks be used in to evaluate promising rice
cultivars in addition to side-by-side comparisons to maintain
perspective of the absolute yield gains being made.

@@

<<fig=TRUE, echo=FALSE, width = 6, height = 5>>=

@@

% \begin{figure}[H]
% <<fig=TRUE, echo=FALSE>>=
% tars <- c("M202","M206")
% comp_yrs = 2004:2005

>>>>

Constraints on arable land are increasing simultaneous with the need
to increase total food production to meet a growing demand
\citep{Foley2011a, Godfray2010, Mueller2012}, which has necessitated
harvesting more grain per unit land area \citep{Lobell2009,
Tittonell2014}. Historically, agricultural research has been
successful in staving off the ``Malthusian catastrophe'' of demand
surpassing supply via continued yield improvement of staple
crops. However, many production systems are experiencing plateaus in
grain yield \citep{Grassini2013, Peng2010}. If this trend continues,
improvements per unit land area are no longer possible, and an
increase in the area under cultivation will be needed to meet food
demand, which carries with it undesirable ecological
implications~\citep{Foley2011a, tilman1999}. Therefore, it is of
critical importance to better understand why the rate of increase in
grain yields has declined or leveled off in intensified production
systems.

California (CA) rice represents one such production system. Rice is
grown primarily in the Sacramento Valley which is characterized by
having a Mediterranean climate, with long days, dry growing season,
and relatively free of pests and diseases, which lead to some of the
highest yields in the world~\citep{FAO2016}. Production in CA is
predominately focused on premium quality medium grain
\textit{japonica} cultivars (e.g., CalRose rice), and rice from CA is
recognized globally for its quality
(\url{http://agfax.com/2015/11/03/rice-calrose-wins-best-rice-world-competition/}). Most
cultivars in use in CA are developed by the California Cooperative
Rice Research Foundation (CCRRF; a collaboration between the
University of California, the USDA-Agricultural Research Service, and
farmer-funded research). In part due to improvements in rice genetics,
rice yields in CA increased rapidly during the period from 1920 --
1990; however, since the 1990s, the rate of yield increase has slowed
(Fig. \ref{fig:state_yield}) despite continuous crop
improvement. Based on a yield-gap analysis, on-farm yields in the
major rice growing region of CA is 73 to 76\% of maximum yield
potential~\citep{espe16a}. In highly intensive systems such as this,
\citet{grassini2011} have shown that farmers are capable of attaining
85\% of the maximum yield potential. Thus, given that farmers are near
the attainable yield potential, increases in yield are likely to be
slow and \citet{espe16a} reported that indeed this was the case
with yields on average increasing by about 50 to 62 kg ha\tsups{-1} between
1999 and 2014.

Certainly one question in looking at this situation is, ``do new
varieties have increased yield potential''?  Like many breeding
programs, one of the goals of the CCRRF is to increase yields through
breeding varieties with increased yield potential. However, when the
popular cultivar M-202 was released in 1985, it was reported to yield
an average of 11 Mg ha\tsups{-1} \citep[14\% moisture content
(MC);][]{m202}. In 2015, the reported average yield of the newest
released cultivar M-209 was 10.8 Mg ha\tsups{-1} \citep[14\%
MC;][]{m209}. However, in annual breeding reports and varietal release
announcements, based on yield performance in side-by-side variety
trails, data show that new varieties usually out yield the older
varieties. In other rice systems, a decrease in a cultivar's yield
performance over time has been observed \citep{Peng1998, Peng2010,
Peng2000, Datta1995}, often termed yield erosion \citep{Peng1998}. It
has been speculated that this yield erosion is due to the inability of
a cultivar to adapt to changing biotic and abiotic
conditions~\citep{Peng1998, Peng2010, Peng2000, Datta1995}. Yield
erosion may explain the apparent contradiction above, in which
side-by-side yield comparisons show yield improvements in varieties
while overall there may be little change in yield potential. To
address this issue we tested the hypothesis that if rice cultivar
yields erode over time then apparent yield increases in side-by-side
yield comparison tests will not reflect true yield increases.o

@@

As part of the annual Statewide Variety Trials, newly developed
entries are tested against officially released cultivars (checks) at
the RES in several trials spanning multiple planting dates per
year. For the purposes of this study, we concentrated on released
medium grain cultivars, as these cultivars constitute approximately
90\% of the planted rice area in CA. Newly developed entries are typically
evaluated in these trials for 3 to 5 years prior to
their official release. For these cultivars planted in breeding trials
at the RES from \Sexpr{min(mod_data[,'year'])} --
\Sexpr{max(mod_data[,'year'])} plot level observations (replicated
3--4 times in a CRB design) were collected leading to a final data set
of \Sexpr{format(nrow(mod_data), big.mark = ',')} observations
representing \Sexpr{length(unique(mod_data[,'id']))} cultivars and
included every publicly released medium grain cultivar developed by
the CCRRF from \Sexpr{min(mod_data[,'release_yr'])} --
\Sexpr{max(mod_data[,'release_yr'])}. For each cultivar, the official
year of public release and the time elapsed (in years) since release was
determined.

@@

\subsection{Cultivar yields at time of release}

There was agreement between the estimates from the Bayesian and
classical approaches.  The Bayesian estimates are presented here,
though the classical estimates can be found in the supplemental
material.  Comparing the cultivars' yield during each's initial time
(average 3 years) in the experimental trials just before being
released showed that across cultivars, yields ranged from
\Sexpr{round(plot_data["M207", "mean"], 2)} (M-207) to
\Sexpr{round(plot_data["M205", "mean"], 2)} Mg ha\tsups{-1}
(M-205). Evaluating these initial yields over time indicates there was
no evidence of genetic yield improvements during the 30-year study
period (Fig. 2). Individual exceptions to this were two varieties
(M-204 and M-205) that had higher yields than the others.

@
\caption{The relationship between yield and the years since release
  for medium grain rice cultivars. Negative values on the X axis
  indicate experimental yields prior to the official release. The
  estimated erosion of yield over time (averaged over differences in
  cultivar and year) is estimated to be \Sexpr{yrs_coef} kg ha\tsups{-1}
  year\tsups{-1}. Points ($\bullet$) and bars (|) represent the mean and
  standard error of the raw data, respectively. The shaded area is the
  90\% credible interval.}
\label{fig:RES_year}
\end{figure}

The yields of each cultivar were evaluated over time relative to the
time when the cultivar was released using the full model which
accounts for both yield erosion and yield improvements over time. The
average decrease in yield over time (i.e., yield erosion) once a
cultivar was selected was estimated \Sexpr{yrs_coef} kg ha\tsups{-1}
year\tsups{-1} (90\% CI: \Sexpr{RES_CI[2]} to \Sexpr{RES_CI[1]}) (Fig
\ref{fig:RES_year}; Table \ref{tbl:coef}). The oldest cultivar in the
trials (\Sexpr{unique(mod_data[which.min(mod_data[,"year"]),"id"])}
had lost an average of \Sexpr{round(yrs_coef*-36/1000,1)} Mg
ha\tsups{-1} since its release. The most dominate medium grain rice
cultivar, M-202 (widely planted from the late 1980s until the early
2000s) experienced similar trends in yield decline
(Fig. \ref{fig:RES_year}; Table \ref{tbl:coef}). M-206 replaced M-202
as the most widely grown cultivar in CA and the current estimated 560
kg ha-1 yield advantage of M-206 over M-202 can be attributed solely
to the erosion of M-202's yield over time (Figs \ref{fig:first_yrs} \&
\ref{fig:compare_yrs}). Further supporting this estimate of yield
erosion, both cultivars yielded similarly at their respective year of
release (Fig. \ref{fig:compare_yrs}), despite the fact that M-202 was
released 19 years prior to M-206 (Fig. \ref{fig:compare_yrs}, Table
\ref{tbl:ids}).

When yield erosion was not corrected for in the analysis (i.e.,
omitted from the model), there is an apparent small increase per year
(approximately \Sexpr{only_release_coef} kg ha\tsups{-1}
year\tsups{-1}; 90\% CI \Sexpr{only_release_CI[1]} to
\Sexpr{only_release_CI[2]}) (Fig. \ref{fig:wo_erosion}).  However,
using the full model which corrected for both the year of release and
yield erosion, indicates that the estimated change in grain yield per
year in the CCRRF's breeding trials at the RES was
\Sexpr{release_yr_coef} kg ha\tsups{-1} year\tsups{-1}, with both
small positive and negative trends within the credible interval (90\%
CI: \Sexpr{rel_CI[1]} to \Sexpr{rel_CI[2]})
(Fig. \ref{fig:first_yrs}).

After correcting for yield erosion, some cultivars were higher
yielding compared to the overall average (as estimated by random
intercepts). The marginal estimate for the performance of a cultivar
averaged over the uncertainty in the other parameters in the model
(i.e., the random effect) shows three cultivars with higher than
average yields:
\Sexpr{paste(gsub('M','M-',rev(row.names(id_est))[c(1,2,4)]),collapse=',')}
(Figs. \ref{fig:first_yrs} \& \ref{fig:ids_comp}).

@@

@@

In California, the cultivars with broad adaptation and adoption over
the 30+ year period of this study are M-202, M-206, and M-105. Cold
tolerance is a major problem for much of the California rice producing
region ~\citep{espe2017}, although it is a bigger problem in the
southern portion of the Sacramento Valley due to the influence of the
ocean winds through the Delta ~\citep{espe16a}. These three
cultivars have a higher degree of cold tolerance, high yield
potential, and relatively short duration and are thus broadly grown
with over 70\% of the area in one of these cultivars during the study
period ~citep{ucce}. Importantly, not all medium grain cultivars that were
released were purported to have higher yields than previous
cultivars. Some cultivars had special traits, but upon release had
similar yields to more broadly adapted cultivars. Examples of this
include M-208, which had resistance to blast disease (scientific name)
and M-402 which is a premium quality variety. Other varieties such as
M-205 and M-209, while high yielding, do not have the cold tolerance
of the other cultivars and are thus only recommended for the northern
portion of the Sacramento Valley.

Looking at all the cultivars that have been released, there is no
conclusive evidence of increased genetic yield potential in medium
grain rice cultivars released by CCRRF's breeding improvement program
over the period \Sexpr{min(mod_data[,'year'])} to
\Sexpr{max(mod_data[,'year'])}, which has stayed relatively constant
at \Sexpr{round(mean(mod_data[,"yield_kg"]), 2)} Mg
ha\tsups{-1}. Since the estimated yield improvement spans from
\Sexpr{rel_CI[1]} to \Sexpr{rel_CI[2]} kg ha\tsups{-1} year\tsups{-1}
(Fig. \ref{fig:RES_year}; Table \ref{tbl:coef}),
both increases and decreases in the genetic yield potential over the
study period are plausible. The estimated yield
improvement of \Sexpr{only_release_coef} kg ha\tsups{-1}
year\tsups{-1} (Fig. \ref{fig:wo_erosion}) that does not account for
yield erosion is lower than the 37 to 60 kg ha\tsups{-1}
year\tsups{-1} estimated by~\citet{Omar2014} in side-by-side
comparisons. Rather, this analysis suggests that the CCRRF breeding
program has maintained but not substantially improved the yield
potential in the face of changing biotic and abiotic pressures. This
conclusion holds true even when evaluating the cultivars which have
been breed for broad adaption and high yield potential (M-202, M-206
and M-105). Yields of these varieties at time of release have all
yielded between
\Sexpr{tmp=mod_data[mod_data[,"yrs_in_trial"]<=0&mod_data[,"id"]%in%c("M202","M206","M105"),];round(min(by(tmp,tmp[,"id"],function(x)mean(x[,"yield_kg"]))/1000),2)}
and
\Sexpr{round(min(by(tmp,tmp[,"id"],function(x)mean(x[,"yield_kg"]))/1000),2)}
Mg ha\tsups{-1} with no indication of yields increasing
(Fig. \ref{fig:first_yrs}). While these yields have not increased,
M-205 did have higher yield potential; however, this variety is not
broadly adapted due to poor cold tolerance.

Yields of a particular cultivar may decline overtime for a number of
reasons. First, ~\citet{Nie2009} reported yield declines in rice due
to continuously flooded conditions over a long time period which could
be overcome with improved N management. This is likely not the case
here as the fields used in this study grow rice only once season a
year and tend to be non-flooded for the remainder of the
year. Furthermore, at this location and unlike most CA rice fields,
the fields are left fallow every other year. Second, changing climate
may influence yields with either warmer~\citep{van2013} or
cooler (Espe et al., 2017) temperatures reducing yields. Again, this
is unlikely to be the case in this situation as Espe et al. (2017)
found no significant changes in growing season temperatures (either
high, low or average temps) the Sacramento Valley during the study
period.  Thirdly, yield erosion of a cultivar as has been reported for
the cultivar IR8 at IRRI~\citep{Peng1998, Peng2000, Peng2010}. Biotic
and abiotic forces have previously been speculated to be responsible
for yield erosion, since the cultivar's genetics remain stable from
the time of selection but the conditions the cultivar may experience
over time are continuously changing~\citep{Peng1998, Peng2000,
Peng2010}. We estimate that yield erosion has decreased the yield of
medium grain rice cultivars released by the CCRRF over the last 30+
years by an average of \Sexpr{yrs_coef} Mg ha\tsups{-1}, potentially
offsetting any gains made by the crop improvement program. However, it
is not clear what is responsible for this erosion. It is unlikely
biotic factors as pest and disease pressure are low in this
Mediterranean climate. Furthermore, these fields are fallowed every
other year and the residues are burned, further lowering the potential
for pest and disease pressure. While efforts should be made to
identify the cause of erosion in this environment, this observation
highlights both the need for maintenance of plant breeding and the
difficulty of increasing genetic yield potential, especially in
systems such as CA where yield is approaching the physiological limit
\citep{espe16a}. Without maintenance breeding (i.e., continuously
selecting cultivars well adapted to current conditions), CA rice
production systems are predicted to have lost an estimated
\Sexpr{round(yrs_coef*(-36/1000),1)} Mg ha\tsups{-1} grain yield over
the study period.

A goal of many breeding programs is to not only maintain the benchmark
yield of the previous ``standard'' but to surpass it, especially if we
are to meet the target of 50\% increased food production by 2050
\citep{Foley2011a}. Given the estimated yield erosion, to hit an
ambitious target of 0.5 Mg ha\tsups{-1} (approximately 5\%) increase
in yield per release (with roughly 4 years between releases), a new
cultivar needs to yield an additional \Sexpr{round(yrs_coef*-4,-1)} kg
ha\tsups{-1} in side-by-side comparison in addition to the 0.5 Mg
ha\tsups{-1} target improvement. Often side-by-side yield improvements
show an improvement in grain yield of 2\% per release, in which case
no absolute improvements in yield have been made beyond maintaining
yield. The evidence here is consistent with the observation that the
genetic yield potential has not increased substantially in the last
30+ years (Fig. \ref{fig:first_yrs}).

@@

There are several cultivars that were found to have better than
average yields, even after correcting for yield erosion. However,
these cultivars (M-201, M-204, M-205, and possibly M-209;
Fig. \ref{fig:ids_comp})) are not widely adopted. It is known that
cool temperatures during the booting stage can have a large impact on
grain yields \citep{espe2017}, and the RES
experiences warmer temperatures than many other rice growing areas in
CA \citep{espe2017}. These cultivars may perform well at the RES but
not widely in CA. Therefore, it should be noted that while our
investigation has been concerned about increasing yield potential, the
estimates here might not capture yield stability. Yield stability is
notoriously difficult to quantify \citep[e.g.,][]{lobell2011}, hence
further investigation is needed to explore how yield stability has
changed over the study period.

@@

California state average yields have increased slowly
despite there being little to no gains in the breeding trials
(Fig. \ref{fig:state_yield}). The evidence here suggests mechanisms
other than increased genetic yield potential are responsible for these gains
(Fig. \ref{fig:compare_yrs}). Improvement in agronomic management,
including more efficient nutrient management \citep{Lundy2012,
Lundy2015, linquist2009}, weed management \citep{Saito2008,
Pittelkow2012, linquist2008}, and other factors may all be contributing to
increases in the state average yield. However, these gains are far less
than those experienced during the preceding era, and by their
nature agronomic gains cannot surpass the genetic yield potential
\citep{Fischer2015}. Judging from yields achieved at the RES
(Figs. \ref{fig:RES_year} \& \ref{fig:compare_yrs}), a
researcher-managed production system implementing best management
practices in more controlled situations compared to farmers' fields, state
average yields (approximately 10 Mg ha\tsups{-1} at the time of
writing; Fig. \ref{fig:state_yield}) may be approaching the current
genetic limit. Assuming that the level of management possible in small
experimental plots cannot be attained in farmers' fields and hence
experimental yields represent the attainable yield ceiling
\citep{Fischer2015}, agronomic gains in farmer's fields are nearing
their potential, consistent with physiological model-based estimates
of the attainable yield gap in CA rice production systems
\citep{espe16a}.

Planning to meet future
demand necessitates increases in grain yield \citep{Godfray2010,
Foley2011a, Mueller2012}. Meeting the challenge to produce more grain
without expansion of land under cultivation will rely on new genetics
with increased yield potential. The evidence here suggests that to
accomplish true increases over time, the improvement target needs take
into account erosion in the current ``standard'' cultivars' yield
over time (Fig. \ref{fig:theory}). While side-by-side comparisons will
continue to be useful, comparisons across years to each cultivar's
benchmark yield at the time of release should provide information on
the absolute improvements being made in yield potential over time. New
technologies may need to be adapted to further increase yield
potential, including hybrid rice.

@@

Here we show that side-by-side cultivar comparisons do not necessarily
indicate gains in yield potential as they do not account for yield
erosion. Quantifying changes in yield potential requires also
examining historical benchmarks. Improvement in grain yield is
essential to meeting the challenges of feeding a growing population in
a changing world. To achieve these improvements, we discuss here that
(1) continuous maintenance via plant improvement is required, without
which yields would decrease over time, (2) the challenge of improving
genetic yield potential is substantial, since absolute gains require
improvement in yield in excess of the maintenance threshold, and (3)
increasing yield through improved management (agronomy) cannot exceed
the ceiling of genetic yield potential. Taken together, these
conclusions underscore both the necessity and substantial challenges
facing rice breeding. We suggest here that historical benchmarks be
used in to evaluate promising rice cultivars in addition to
side-by-side comparisons to maintain perspective of the absolute yield
gains being made.

@@

<<fig=TRUE, echo=FALSE, width = 6, height = 5>>=
tars <- c("M202","M206")

@@

@
\caption{The performance of two major cultivars over the period
\Sexpr{min(mod_data[,'year'])} -- \Sexpr{max(mod_data[,'year'])}. The
solid line is the estimated yield erosion (\Sexpr{yrs_coef} kg
ha\tsups{-1} year\tsups{-1}), while the dashed line is the published
3-year average yield of M-202 at its release (1982-1984). Both
cultivars have a similar yield at their respective year of release,
and similar decreases over time.}
\label{fig:compare_yrs}
\end{figure}
\newpage
% \begin{figure}[H]
% <<fig=TRUE, echo=FALSE>>=
% tars <- c("M202","M206")
% comp_yrs = 2004:2005
